---
title: "Delta Data Presentation"
author: "Barney Ricca"
date: "12/16/2017"
output: word_document
---
To do, based on our conversations:



###Graphs of Robust Effect Size Estimates
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

c("data.table", "plyr", "dplyr", "ggplot2",
  "here", "lawstat", "sjmisc", "tidyr") -> package_names
for(package_name in package_names) {
  if(!is.element(package_name, installed.packages()[,1])) {
     install.packages(package_name, repos="http://lib.stat.cmu.edu/R/CRAN")
  }
  library(package_name, character.only=TRUE,quietly=TRUE,verbose=FALSE)
}
rm(list=c("package_names","package_name"))

sessionInfo()  # Information about all the packages and versions

Sys.time() # So we have a record of when this was knit
```

```{r functionDefs}
# Functions listed in alphabetical order

b_delta <- function(a1, a2) {
  length(a1) -> l1
  length(a2) -> l2
  return(abs((median(a1)-median(a2))) / 
           (((l1-1)*mad(a1)+(l2-1)*mad(a2))/(l1+l2-2)))
}

c_d <- function(a1, a2) {
  # a1 and a2 are two samples (of equal sizes?)
  # 2017-10-17: Corrected the pooling for equal sample sizes from a denominator of 2 to a denominator of sqrt(2). If, however, sd(a1)==sd(a2), then it should simply be sd(a1).
  if(sd(a1)!=sd(a2)) {
    return(abs((mean(a1)-mean(a2)))/(sqrt(var(a1)+var(a2))/sqrt(2)))
  } else
  {
    return(abs((mean(a1)-mean(a2)))/sd(a1))
  }
}

cnorm <- function(n, mean=.5, sd1=1, sd2=5, prob=0.2) {
  # creating a contaminated normal distribution
  # n: number of samples to return
  sd <- sample(c(sd1,sd2), n, replace=T,
                prob=c(1-prob,prob))
  return (rnorm(n, mean, sd))
}

contamtest <- function(N=10000, n=250, p=0.2, ci=0.95, sigma=1) {
  # A function to facilitate the testing of N, p, and n
  # N = number of samples to draw
  # n = size of sample
  # p = contamination probability
  # ci = confidence interval band (e.g., 95% is the default)

  contam=cnorm(1e6, prob=p) # contaminated normal to sample from

  replicate(N,rnorm(n,0,sigma)) -> a1
  replicate(N,sample(contam,n,replace=TRUE)) -> a2

  # a1 and a2 are matrices with N columns and n rows.
  vector(mode="numeric", length=N) -> cohensd
  vector(mode="numeric", length=N) -> blainesdelta
  for(i in 1:N) {
    c_d(a1[,i], a2[,i]) -> cohensd[i]
    b_delta(a1[,i], a2[,i]) -> blainesdelta[i]
  }

  # To return: a data frame
  (1-ci)/2 -> q # Lower quantile
  data.frame("d" = mean(cohensd),
             "delta" = median(blainesdelta),
             "d.lwrCI" = quantile(x = cohensd, probs = q),
             "d.uprCI" = quantile(x = cohensd, probs = 1-q),
             "delta.lwrCI" = quantile(x = blainesdelta, probs = q),
             "delta.uprCI" = quantile(x = blainesdelta, probs = 1-q)) -> ret.df
  return(ret.df)
}

c2test <- function(N=10000, n=250, ci=0.95) {
  # Nota bene: The old version of this function would find the CI of the
  #  mean of _d_ and $\delta$ rather than the CI of _d_ and $\delta$.
  # A function to facilitate the testing of N, p, and n
  # N = number of samples to draw
  # n = size of sample
  # p = contamination probability
  # ci = confidence interval band (e.g., 95% is the default)

  replicate(N,rnorm(n,0,1)) -> a1
  replicate(N,rnorm(n,0.5,5)) -> a2

  # a1 and a2 are matrices with N columns and n rows.
  vector(mode="numeric", length=N) -> cohensd
  vector(mode="numeric", length=N) -> blainesdelta
  for(i in 1:N) {
    c_d(a1[,i], a2[,i]) -> cohensd[i]
    b_delta(a1[,i], a2[,i]) -> blainesdelta[i]
  }

  # To return: a data frame
  (1-ci)/2 -> q # Lower quantile
  data.frame("d" = mean(cohensd),
             "delta" = median(blainesdelta),
             "d.lwrCI" = quantile(x = cohensd, probs = q),
             "d.uprCI" = quantile(x = cohensd, probs = 1-q),
             "delta.lwrCI" = quantile(x = blainesdelta, probs = q),
             "delta.uprCI" = quantile(x = blainesdelta, probs = 1-q)) -> ret.df
  return(ret.df)
}

deltatest <- function(N=10000, n=250, ci=0.95) {
  # A function to facilitate the testing of N and n
  # N = number of samples to draw
  # n = size of sample
  # ci = confidence interval band (e.g., 95% is the default)

  replicate(N,rnorm(n,0,1)) -> a1
  replicate(N,rnorm(n,0.5,1)) -> a2  # sd varies here

  # a1 and a2 are matrices with N columns and n rows.
  vector(mode="numeric", length=N) -> cohensd
  vector(mode="numeric", length=N) -> blainesdelta
  for(i in 1:N) {
    c_d(a1[,i], a2[,i]) -> cohensd[i]
    b_delta(a1[,i], a2[,i]) -> blainesdelta[i]
  }

  # To return: a data frame
  (1-ci)/2 -> q # Lower quantile
  data.frame("d" = mean(cohensd),
             "delta" = median(blainesdelta),
             "d.lwrCI" = quantile(x = cohensd, probs = q),
             "d.uprCI" = quantile(x = cohensd, probs = 1-q),
             "delta.lwrCI" = quantile(x = blainesdelta, probs = q),
             "delta.uprCI" = quantile(x = blainesdelta, probs = 1-q)) -> ret.df
  return(ret.df)
}

displaycomp <- function(N, n, df) {
  cat("N = ",N,"  n = ",n,'\n')
  cat("mean(Cohen's d): ",'\t', df$d,'\n')
  cat("median(delta): ",'\t',df$delta,'\n')
  cat('\n')
  cat("95% CI of d: ",'\t', '\t', df$d.lwrCI,'\t',
      df$d.uprCI,'\n')
  cat("95% CI of delta: ",'\t',
      df$delta.lwrCI,'\t',df$delta.uprCI,'\n')
  cat('\n')
}
```

### Results

#### Load Data
```{r loadData}
# Data from rnorm(0,1) and rnorm(0.5,1)
load("sample size - homoscedastic.RData")  # delta.df
delta.df -> overlap_equal.df
colnames(overlap_equal.df)[1] <- "n"

# Data from rnorm(0,1) and rnorm(5,1)
load("sample size - homoscedastic - far.RData")
delta.df -> separate_equal.df
colnames(separate_equal.df)[1] <- "n"

# Data from rnorm(0,1) and rnorm(0.5,1), with varying N
load("homoscedastic - replications.RData")
delta.df -> replicate.df

# Many different unequal variances: rnorm(0,1) vs. rnorm(0.5, sigma)
load("Unequal variances.RData")
ratio.df -> unequal.df

# First contamination: rnorm(0,1) vs. p*rnorm(0.5,1) U (1-p)*rnorm(0.5,5)
# Forgot to copy
load("Contaminated normals.RData")
effsize.df -> contam1.df

load("Contaminated narrow normals.RData")
effsize.df -> contam2.df

# Second contamination: upper contaminated by (2,4)
load("upper contamination.RData")
results.df -> upper_contam.df

# Second contamination: upper contaminated by (-3,-1)
load("lower contamination.RData")
results.df -> lower_contam.df

# rnorm(5,1) contaminated by (7,9)
load("separated upper contamination.RData")
results.df -> sep_upper_contam.df

# rnorm(.5,1) contam by (1.5, 3.5)
load("separated lower contamination.RData")
results.df -> con_upper_contam.df

# rnorm(0.5,1) contaminated by rnorm(3,1)
load("separated normal contamiatnion.RData")
results.df -> norm_norm.df

```

### The plots
What needs to be plotted for results?

#### Impact of n: rnorm(0,1) and rnorm(0.5, 1)
This doesn't take long to run, so do not create the intermediate data frame.
```{r}
overlap_equal.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., n!=1) %>%
  ggplot(., aes(x=n, y=EffectSize, shape = Measure)) +
  scale_shape_manual(values = c(1,15)) + 
  theme_bw() +
  xlab("Sample size") + 
  ylab("Effect size") +
  geom_point() -> g
g + ggtitle("Effect Size vs. Sample Size",
            subtitle = "Two normal populations; mean difference = 0.5 * standard deviation") +
  geom_smooth(method = "loess", se = FALSE, 
              aes(linetype = Measure), color = 'black',
              show.legend = TRUE) +
  ylim(0.495, 0.53)
ggsave("1a.png")

g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) +
ge
ggsave("1b.png")
```

#### Impact of n: rnorm(0,1) and rnorm(5,1)
```{r}
separate_equal.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., n!=1) %>%
  ggplot(., aes(x=n, y=EffectSize, colour=Measure)) +
  theme_bw() +
  geom_point() -> g
g + ggtitle("Effect Size vs. Sample Size",
            subtitle = "Two normal populations; mean difference = 5 * standard devaition")
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge + ggtitle("Effect Size vs. Sample Size",
             subtitle = "with 95% confidence intervals")

```

#### Impact of N: Replcations
```{r}
replicate.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  ggplot(., aes(x=N, y=EffectSize, colour=Measure)) +
  theme_bw() +
  geom_point() -> g
g + ggtitle("Effect Size vs. Number of Replications")
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge + ggtitle("Effect Size Confidence Intervals vs. Number of Replications")

```


#### Impact of sigma: Heteroscedastic: Impact of sigma2/sigma1
```{r}
# Do a middle sample size for effect size by sigma ratio
unequal.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., n==100) %>%
  ggplot(., aes(x=r, y=EffectSize, colour=Measure)) +
  theme_bw() +
  ggtitle("Effect Size vs. Standard Deviation ratio", 
          subtitle = "sample size = 100") +
  xlab("Ratio of Standard Deviations") + 
  geom_point() -> g
g
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge

# Do a small sample size for effect size by sigma ratio
unequal.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., n==10) %>%
  ggplot(., aes(x=r, y=EffectSize, colour=Measure)) +
  theme_bw() +
  ggtitle("Effect Size vs. Standard Deviation ratio", 
          subtitle = "sample size = 10") +
  xlab("Ratio of Standard Deviations") + 
  geom_point() -> g
g
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge

# Do a large sample size for effect size by sigma ratio
unequal.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., n==250) %>%
  ggplot(., aes(x=r, y=EffectSize, colour=Measure)) +
  theme_bw() +
  ggtitle("Effect Size vs. Standard Deviation ratio", 
          subtitle = "sample size = 250") +
  xlab("Ratio of Standard Deviations") + 
  geom_point() -> g
g
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge

```

#### Impact of p: Probability contamination
```{r}

contam1.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., Sample==250) %>%
  ggplot(., aes(x=Prob, y=EffectSize, shape=Measure)) +
  theme_bw() +
  xlim(0,0.5) +
  scale_shape_manual(values = c(1,15)) + 
  geom_smooth(method = "loess", aes(linetype = Measure)) +
  xlab("Fraction of points contaminated") +
  ylab("Effect size") +
#  ggtitle("Effect Size vs. Standard Deviation ratio", 
#          subtitle = "sample size = 250") +
#  xlab("Ratio of Standard Deviations") + 
  geom_point() -> g
g
ggsave("2.png")
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge

```
```{r}
contam2.df %>%
  to_long(., keys="Measure",
          values = c("EffectSize","LowerCI","UpperCI"),
          c("delta","d"), c("delta.lwrCI","d.lwrCI"),
          c("delta.uprCI","d.uprCI")) %>%
  filter(., Sample==250) %>%
  ggplot(., aes(x=Prob, y=EffectSize, shape=Measure)) +
  theme_bw() +
  xlim(0,0.5) +
  scale_shape_manual(values = c(1,15)) + 
  geom_smooth(method = "loess", aes(linetype = Measure)) +
  xlab("Fraction of points contaminated") +
  ylab("Effect size") +
#  ggtitle("Effect Size vs. Standard Deviation ratio", 
#          subtitle = "sample size = 250") +
#  xlab("Ratio of Standard Deviations") + 
  geom_point() -> g
g
#ggsave("2.png")
g +
  geom_errorbar(aes(ymin=LowerCI, ymax=UpperCI, group=Measure)) -> ge
ge

```


#### Impact of b: rnorm(0.5,1) contaminated by runif(2,4)
```{r}
upper_contam.df %>% filter(., Levene != 0.1 & Delta != 0.1) -> upper_contam.df

upper_contam.df %>%
  filter(., Shapiro >= 0.1) %>%
  filter(., Levene >= 0.1) %>%
  filter(., as.numeric(Replacements) < 25) %>% 
  group_by(., Replacements) %>%
  dplyr::summarise(., "Delta"=median(Delta), "d"=mean(d), "Count"=n()) %>%
  gather(Measure, EffectSize, -Replacements, -Count) %>%
  ggplot(.,
         aes(x=as.numeric(Replacements), y=EffectSize, 
             shape=Measure, size=Count)) +
  theme_bw() +
  scale_shape_manual(values = c(1,15)) + 
#  ggtitle("Effect Size vs. Number of Replacements", 
#          subtitle = "Total sample size = 100\nBoth Shapiro-Wilk and Levene are significant (p<0.1)\n10000 replications at each replacement level") +
  xlab("Number of Replacements") + 
  ylab("Effect size")
  geom_point() -> g
g

upper_contam.df %>%
  filter(., Shapiro < 0.1) %>%
  filter(., Levene < 0.1) %>%
  filter(., as.numeric(Replacements) < 25) %>% 
  group_by(., Replacements) %>%
  dplyr::summarise(., "Delta"=median(Delta), "d"=mean(d), "Count"=n()) %>%
  gather(Measure, EffectSize, -Replacements, -Count) %>%
  ggplot(.,
         aes(x=as.numeric(Replacements), y=EffectSize, 
             shape=Measure, size=Count)) +
  theme_bw() +
#  ggtitle("Effect Size vs. Number of Replacements", 
#          subtitle = "Total sample size = 100\nBoth Shapiro-Wilk and Levene are not significant (p>0.1)\n10000 replications at each replacement level") +
  xlab("Percent of Replacements") + 
  ylab("Effect size") +
  geom_point() -> gs
gs
ggsave("3.png")
```

#### Impact of b: rnorm(5,1) contaminated by runif(7,9)
Remember:
$H_0$ of Levene is homoscedastic
$H_0$ of Shapiro is normal
```{r}
sep_upper_contam.df %>%
  filter(., Shapiro >= 0.1) %>%
  filter(., Levene >= 0.1) %>%
  filter(., as.numeric(Replacements) < 50) %>% 
  group_by(., Replacements) %>%
  dplyr::summarise(., "Delta"=median(Delta), "d"=mean(d), "Count"=n()) %>%
  gather(Measure, EffectSize, -Replacements, -Count) %>%
  ggplot(.,
         aes(x=as.numeric(Replacements), y=EffectSize, 
             color=Measure, size=Count)) +
  theme_bw() +
  ggtitle("Effect Size vs. Number of Replacements", 
          subtitle = "Total sample size = 100\nBoth Shapiro-Wilk and Levene are significant\nNumber of replications per replacement level=100") +
  xlab("Number of Replacements") + 
  geom_point() -> g
g
```

#### Still to do:

Impact of b: rnorm(5,1) contaminated by runif(-3, -1) 
```{r}
lower_contam.df %>% filter(., Levene != 0.1 & Delta != 0.1) -> lower_contam.df

lower_contam.df %>%
  filter(., Shapiro >= 0.1) %>%
  filter(., Levene >= 0.1) %>%
  filter(., as.numeric(Replacements) < 25) %>% 
  group_by(., Replacements) %>%
  dplyr::summarise(., "Delta"=median(Delta), "d"=mean(d), "Count"=n()) %>%
  gather(Measure, EffectSize, -Replacements, -Count) %>%
  ggplot(.,
         aes(x=as.numeric(Replacements), y=EffectSize, 
             shape=Measure, size=Count)) +
  theme_bw() +
  scale_shape_manual(values = c(1,15)) + 
#  ggtitle("Effect Size vs. Number of Replacements", 
#          subtitle = "Total sample size = 100\nBoth Shapiro-Wilk and Levene are significant (p<0.1)\n10000 replications at each replacement level") +
  xlab("Number of Replacements") + 
  ylab("Effect size") +
  geom_point() -> g
g

lower_contam.df %>%
  filter(., Shapiro < 0.1) %>%
  filter(., Levene < 0.1) %>%
  filter(., as.numeric(Replacements) < 25) %>% 
  group_by(., Replacements) %>%
  dplyr::summarise(., "Delta"=median(Delta), "d"=mean(d), "Count"=n()) %>%
  gather(Measure, EffectSize, -Replacements, -Count) %>%
  ggplot(.,
         aes(x=as.numeric(Replacements), y=EffectSize, 
             shape=Measure, size=Count)) +
  theme_bw() +
#  ggtitle("Effect Size vs. Number of Replacements", 
#          subtitle = "Total sample size = 100\nBoth Shapiro-Wilk and Levene are not significant (p>0.1)\n10000 replications at each replacement level") +
  scale_shape_manual(values = c(1,15)) + 
  xlab("Percent of Replacements") + 
  ylab("Effect size") +
  geom_point() -> gs
gs
ggsave("4.png")

```



Contamination | Percent not found | Delta mean | Delta sd | d mean | d sd
--------------|-------------------|------------|----------|--------|-----
0             | 2.3               |  0.49      | 0.32     | 0.43   | 0.25


However, we also need similar things for when the contamination is found by the tests. Compare the differences

